\chapter{Making Web Requests with HTTP}

The Hypertext Transfer Protocol (HTTP) is the protocol used for
transmitting hypertext over the World Wide Web. It is the foundation
of any data exchange on the web, and it is a protocol used for
transmitting hypertext requests from clients (like a user's browser)
to servers, which respond with the requested resources.\index{HTTP}

\section{HTTP Requests}

An HTTP request is made up of several components:

\begin{itemize}
    \item \textbf{Method}: The HTTP method, like GET (retrieve data), POST (send data), PUT (update data), DELETE (remove data), and so on.
    \item \textbf{URL}: The URL of the resource to retrieve, send data to, update or delete.
    \item \textbf{Headers}: Additional information about the request or response, like the content type of the body.
    \item \textbf{Body}: The body of the request, used when sending data in POST or PUT requests.
\end{itemize}

One way to think about an HTTP request is to imagine ordering food at
a restaurant. The method is the type of action you want to take, such
as asking to see the menu, placing an order, changing an order, or
canceling it. The URL is like telling the server exactly which
restaurant counter to go to. The headers are like the extra details
you give, such as saying you want vegetarian food or that you would
like water with your meal. The body is the actual order information,
such as the list of food you want to buy.

The method tells the server what kind of action the client wants to
perform. For example, a GET request is used when the client wants to
receive information from the server, such as loading a web page or
asking for a list of items. A POST request is used when the client
wants to send new information to the server, such as creating an
account or submitting a form. A PUT request is used to update
something that already exists, and a DELETE request is used to remove
something.

The URL works like an address. It tells the request where to go. If
the method explains what the client wants to do, the URL explains
where that action should happen. For example, if a web app wants a
list of users, it might send a request to a URL that ends in
\verb|/users|. If it wants information for one specific user, it might
send a request to a URL like \verb|/users/5|.

Headers help the client and server understand each other better. They
carry extra information along with the request or response, such as
what type of data is being sent. This is similar to writing
``Fragile'' on a package or adding a note that says ``Please handle
with care.'' The message is still the same package, but the note helps
explain how it should be handled.

The body is the part of the request that contains the actual data
being sent. For example, when a user fills out a sign-up form, the
name, email address, and password they enter may be placed in the body
of the request. Not every request needs a body. GET requests usually
do not send one, because they are mainly asking to receive information
rather than send it.


\section{Using HTTP with Web-Based APIs}


Software developers often use HTTP to interact with web-based APIs. An
Application Programming Interface (API) is a set of rules that allows
programs to talk to each other. The developer creates the API on the
server and allows the client to talk to it.

When a developer makes a request to an API endpoint, they're asking
the server to either send them some data or receive some data from
them. The response from the server will often be in a format like JSON
or XML, which the developer can then use in their own application.

For example, a developer might make a GET request to
\texttt{https://api.example.com/users} to retrieve a list of all
users. The server would respond with a list of users in a format like
JSON.

An API endpoint is a specific URL that a client can use to communicate
with a server. Each endpoint usually has a clear purpose. One endpoint
might return a list of books, while another might return information
about a single book. In the same way that a restaurant has different
counters for ordering food, paying, or picking up an order, an API can
have different endpoints for different tasks.

When a client sends a request to an API, the server reads the request
and decides how to respond. If the request is valid, the server sends
back a response with the requested data or a message saying the action
worked. If something is wrong, the server sends back a response
explaining that there was a problem. This back-and-forth communication
is one of the main ways web applications work behind the scenes.

For example, imagine a school website that stores a list of students.
A client could send a GET request to an endpoint like \texttt{/students}
to ask for the full list. It could send a POST request to the same
endpoint to add a new student. It could send a PUT request to
\texttt{/students/3} to update the information for one student, or a
DELETE request to \texttt{/students/3} to remove that student from the
list. These requests follow the same HTTP ideas, even if the website
or app itself looks very different on the screen.

Many modern apps work this way. A weather app may send a request to
get the latest forecast. A music app may send a request to get a list
of songs. A shopping app may send a request to place an order. Even
though users may only see buttons, menus, and pictures, HTTP requests
are often moving data in the background.


\section{HTTP Responses}

After a server receives an HTTP request, it sends back an HTTP
response. The response is the server's answer to the client. Just as a
request has several parts, a response also has its own important
parts.

An HTTP response usually includes a status code, headers, and
sometimes a body. The status code tells the client what happened. The
headers give extra information about the response. The body contains
the returned data, if there is any.

A simple way to think about a response is to imagine asking a librarian
for a book. The librarian might say, ``Here is the book you asked
for,'' ``That book is checked out,'' or ``We do not have that
book.'' These are different kinds of responses to the same kind of
request.

Some common HTTP status codes are:

\begin{itemize}
    \item 200: The request worked, and the server is sending back the requested data.
    \item 201: The request worked, and the server created something new.
    \item 400: The request had a problem, such as missing or incorrect information.
    \item 404: The server could not find what was requested.
    \item 500: Something went wrong on the server.
\end{itemize}

These codes help the client understand the result quickly. Instead of
reading a long message first, the client can look at the status code
and get a fast summary of what happened.


\section{Data Formats in Responses}

When a server sends data back to a client, that data needs to be
organized in a way both sides can understand. Two common formats are
JSON and XML. These formats are simply ways of arranging information
so that programs can read it clearly and consistently.

JSON is one of the most common formats used with web APIs. It is
popular because it is lightweight and easy for both humans and
programs to read. A JSON response might include data written as
key--value pairs. For example, a user might be described with data
such as a name, age, and email address.

Here is a small example of JSON:

\begin{verbatim}
{
  "name": "Ava",
  "grade": 10,
  "club": "Robotics"
}
\end{verbatim}

In this example, \texttt{"name"}, \texttt{"grade"}, and \texttt{"club"}
are the labels, and \texttt{"Ava"}, \texttt{10}, and
\texttt{"Robotics"} are the values. This is similar to filling out an
information card. Each label tells what kind of information is being
stored, and each value is the actual information.

XML is another format that may be used. It organizes information with
opening and closing tags. While it is less common in many modern
beginner examples, it is still useful to know that different APIs may
return data in different formats.
