\chapter{The Greek Alphabet}

If you do anything involving math or physics, you will be encountering Greek letters on a regular basis. Here is a table for your reference:

\begin{tabular}{c c  l | c c l}
Capital & Lower & Pronounced & Capital & Lower & Pronounced\\
\hline
$A$ & $\alpha$ & Alpha & $N$ & $\nu$ & Nu\\
$B$ & $\beta$ & Beta & $\Xi$ & $\xi$ & Xi ("ku-ZY") \\
$\Gamma$ & $\gamma$ & Gamma & $O$ & $o$ & Omicron\\
$\Delta$ & $\delta$ & Delta & $\Pi$ & $\pi$ & Pi\\
$E$ & $\epsilon$ & Epsilon & $P$ & $\rho$ & Rho\\
$Z$ & $\zeta$ & Zeta & $\Sigma$ & $\sigma$ & Sigma\\
$H$ & $\eta$ & Eta & $T$ & $\tau$ & Tau\\
$\Theta$ & $\theta$ & Theta & $\Upsilon$ & $\upsilon$ & Upsilon\\
$I$ & $\iota$ & Iota & $\Phi$ & $\phi$ & Phi\\
$K$ & $\kappa$ & Kappa & $X$ & $\chi$ & Chi ("Kai")\\
$\Lambda$ & $\lambda$ & Lambda & $\Psi$ & $\psi$ & Psi ("Sigh")\\
$M$ & $\mu$ & Mu & $\Omega$ & $\omega$ & Omega
\end{tabular}

\section{Greek Letters in Math and Science}
These are important to know, for a large variety of reasons. Let's look at the most common physics and math uses:

\begin{tabular}{c l}
Symbol & Common meaning \\
\hline
$\pi$ & circle constant, approximately $3.14159$ \\
$\theta$ & an angle \\
$\Delta$ & change in a quantity \\
$\Sigma$ & sum of many terms \\
$\lambda$ & wavelength \\
$\mu$ & coefficient of friction, mean, or micro- \\
$\rho$ & density \\
$\omega$ & angular velocity \\
$\Omega$ & ohms, a unit of electrical resistance \\
\end{tabular}

You will see a lot of equations that reference Greek letters, using them as \emph{placeholders}, \emph{constants}, or \emph{variables} for an unknown quanity. For example, there area of a circle is given by

$$A = \pi r^2$$

where $r$ is the radius of the circle. In this case, $\pi$ is a constant, and $r$ is a variable. In other cases, you might see the physics equation
$$\Delta x = v \Delta t + \frac{1}{2} a \Delta t^2$$

This is the equation for the change in position under constant acceleration, where $\Delta x$ is the change in position, $v$ is the initial velocity, $a$ is the acceleration, and $\Delta t$ is the change in time. In this case, $\Delta x$, $v$, $a$, and $\Delta t$ are all variables.

One final example that we will explore later is the equation for wavelength in terms of frequency and the speed of light:
$$\lambda = \frac{c}{f}$$
where $\lambda$ is the wavelength, $c$ is the speed of light, and $f$ is the frequency. In this case, $\lambda$ and $f$ are variables, while $c$ is a constant. We can then translate lambda into other equations as needed.

Generally, we will define the Greek letters when providing an equation. However, it is important to be familiar with the Greek alphabet, so that you can understand what the symbols mean when you see them in equations.