\chapter{Matrices and Systems of Linear Equations}

\section{Systems of Linear Equations}
matrix-vector multiplication, representing linear combination
system of linear equations connection
linear combo example
system of equations example (trail mix, be more explicit about eqs and representation)
practice
	use matrix multiplication to find...
	grading question moved here

\section{Trail Mix for Mars}
Let's look at an applied problem. Three astronauts, Pat, Kai, and River, are getting ready for a trip to Mars. NASA food service is preparing trail mix for the voyage, tailored to each astronaut's taste. The chef needs to submit a budget based on the cost of the trail mix for each astronaut. The mix is made up of raisins, almonds, and chocolate.

Pat prefers a raisins:almonds:chocolate ratio of 6:10:4, Kai likes 2:3:15, and River wants 14:1:5. The chef can buy a kg of raisins for \$7.50, a kg of almonds for \$14.75, and a kg of chocolate for \$22.25. Assuming each astronaut will get 20 kg of trail mix, which astronaut will cost more to feed?

First, set up a matrix to represent the raisins:almonds:chocolate ratios. (Conveniently, these ratios already add to 20.)

$$MixRatios = \begin{bmatrix}
6 & 10 & 4  \\
2 & 3 & 15 \\
14 & 1 & 5
\end{bmatrix}$$

Use a vector to represent the cost of each item:
$$IngredientCost = 
\begin{bmatrix}
7.50 \\
14.75 \\
22.25
\end{bmatrix}$$

To find the cost of trail mix for each astronaut, we simply find the dot product between the mix ratios and the ingredient costs to get:

Pat =  \$281.50 \newline
Kai =  \$615.50 \newline
River = \$231.00 \newline


\begin{Exercise}[title={Vector Matrix Multiplication}, label=vector-matrix-multiply01]
Multiply the array $A$ with the vector $v$. Compute this by hand, and make sure to show your computations. 
$$A = \begin{bmatrix}
1 & -2 & 3 & 5  \\
-4 & 2 & 7 & 1 \\
3 & 3  & -9 & 1
\end{bmatrix}$$

$$v = 
\begin{bmatrix}
 2 \\
 2 \\
 6 \\
 -1
 \end{bmatrix}$$
\end{Exercise}

\begin{Answer}[ref=vector-matrix-multiply01]
$$Av = (11 \text{ }37 \text{ } -43)$$
\end{Answer}

\begin{Exercise}[title = {Using Vector Matrix Multiplication}, label = vmm02]
A college professor offers three different methods of determining a student's final grade. In method A, the student's grade is 20\% based on attendance, 50\% homework, 15\% midterm, and 15\% final. This professor knows many students can learn the material without attending every class, so with method B the student's grade is 50\% homework, 20\% midterm, and 30\% final. Last, the professor knows some students don't do the homework but still show they understand the material by doing well on the tests. With method C, a student's grade is 40\% midterm and 60\% final. The professor uses whatever method results in the highest grade to determine each student's final grade. 

Suppose Suzy has attended 65\% of classes, has an average homework grade of 30\%, earned a 80\% on the midterm, and earned a 75\% on the final. What final grade will her professor post?
\end{Exercise}

\begin{Answer}[ref = vmm02]
The different methods can be represented in a matrix:
$$\begin{bmatrix}
0.20 & 0.50 & 0.15 & 0.15\\
0 & 0.50 & 0.20 & 0.30\\
0 & 0 & 0.4 & 0.6
\end{bmatrix}$$

And Suzy's individual grades can be represented by a vector:
$$\begin{bmatrix}
65\\
30\\
80\\
75
\end{bmatrix}
$$

To see the results of the three different methods, we can multiply the matrix and the vector:
$$\begin{bmatrix}
0.20 & 0.50 & 0.15 & 0.15\\
0 & 0.50 & 0.20 & 0.30\\
0 & 0 & 0.4 & 0.6
\end{bmatrix}
\cdot 
\begin{bmatrix}
65\\
30\\
80\\
75
\end{bmatrix} = 
\begin{bmatrix}
0.2(65)+0.5(30)+0.15(80) + 0.15(75)\\
0(65) + 0.5(30) + 0.2(80) + 0.3(75)\\
0(65) + 0(30) + 0.4(80) + 0.6(75)
\end{bmatrix}$$

Which yields:
$$\begin{bmatrix}
51.25\\
53.5\\
77
\end{bmatrix}$$

Since method C yields the highest grade, the professor will post a final grade of 77. 
\end{Answer}

\subsection{Vector-Matrix Multiplication in Python}
Most real-world problems use very large matrices, where it becomes impractical to do calculations by hand. As long as you understand how matrix-vector multiplication is performed, you will be equipped to use a computing language, like Python, to do the calculations for you. 

Create a file called \filename{vectors\_matrices.py} and enter this code:

\begin{Verbatim}
# import the python module that supports matrices
import numpy as np

# create an array
a = np.array([[5, 1, 3, -2], 
              [1, -1, 8, 4], 
              [6, 2, 1, 3]])

# create a vector 
b = np.array([1, 2, 3, -8])

# calculate the dot product
print(a.dot(b))
\end{Verbatim}

When you run it, you should see:
\begin{Verbatim}
[16, 6, 8]
\end{Verbatim}

\section{Where to Learn More}
Watch this video from Khan Academy about matrix-vector products: \url{https://rb.gy/frga5}

