

\chapter{Floating Point Arithmetic}

\section{Why Computers Can't Count Perfectly}
Here is something worth sitting with before you read on: type \texttt{0.1 + 0.2} into a Python shell and press Enter.
You might expect to see \texttt{0.3}.
What you actually see is \texttt{0.30000000000000004}.

That is not a bug in Python.
It is not a rounding mistake you made.
It is a fundamental consequence of how every modern computer stores numbers, and understanding it will make you a much more careful analyst.

\subsection{Counting in Base 10 vs.\ Counting in Base 2}

You have been working in \newterm{base 10}\index{base 10} your whole life.
Every digit in a number like $327$ represents a power of ten: $3 \times 10^2 + 2 \times 10^1 + 7 \times 10^0$.
A computer, at the hardware level, only knows two states: on and off, which we write as $1$ and $0$.
So computers count in \newterm{base 2}\index{base 2}, or \newterm{binary}\index{binary}, where every digit represents a power of two.

The number $13$ in binary, for instance, is $1101$, because
\[
  1 \times 2^3 + 1 \times 2^2 + 0 \times 2^1 + 1 \times 2^0 = 8 + 4 + 0 + 1 = 13.
\]

Whole numbers convert between bases cleanly.
Fractions are where things get interesting.

\subsection{The Fraction Problem}

Consider the decimal fraction $\tfrac{1}{3}$.
In base 10, you cannot write it exactly --- it becomes the infinite repeating decimal $0.333\ldots$
You have probably made peace with this.
But notice: $\tfrac{1}{3}$ cannot be represented with a finite number of decimal digits, so whenever you truncate it you introduce a small error.

The same thing happens in binary, but for a different set of fractions.
In base 2, only fractions whose denominator is a power of 2 can be written exactly.
$\tfrac{1}{2} = 0.1_2$, $\tfrac{1}{4} = 0.01_2$, $\tfrac{3}{4} = 0.11_2$ --- all fine.
But $\tfrac{1}{10}$, the number we write as $0.1$, has no exact finite binary representation.
It expands into an infinitely repeating binary fraction, just as $\tfrac{1}{3}$ does in decimal.

\begin{Exercise}[title={Which fractions survive?}, label={ex:fp-exact-fractions}]
  Decide whether each of the following can be represented exactly in binary.
  Write ``exact'' or ``repeating'' in the second column, and give a brief reason.

  \begin{center}
  \begin{tabular}{lll}
    \hline
    Fraction & Exact or repeating in binary? & Reason \\
    \hline
    $\tfrac{1}{2}$  & \phantom{repeating} & \\[6pt]
    $\tfrac{1}{4}$  & & \\[6pt]
    $\tfrac{1}{5}$  & & \\[6pt]
    $\tfrac{3}{8}$  & & \\[6pt]
    $\tfrac{1}{10}$ & & \\[6pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:fp-exact-fractions}]
  $\tfrac{1}{2}$: exact ($0.1_2$, denominator is a power of 2).
  $\tfrac{1}{4}$: exact ($0.01_2$).
  $\tfrac{1}{5}$: repeating (denominator 5 is not a power of 2; $0.1_2$ repeats as $0.0\overline{0011}_2$).
  $\tfrac{3}{8}$: exact ($0.011_2$, denominator $8 = 2^3$).
  $\tfrac{1}{10}$: repeating (same reason as $\tfrac{1}{5}$; in fact $0.1$ in decimal is $0.0\overline{0011}_2$).
\end{Answer}

\subsection{Finite Memory, Infinite Decimals}

A computer cannot store an infinitely long binary fraction.
It has to stop somewhere.
The hardware allocates a fixed number of binary digits, called \newterm{bits}\index{bits}, and stores the best finite approximation it can.
When the true value cannot be stored exactly, the computer rounds to the nearest representable number.
That rounding is called \newterm{floating-point error}\index{floating-point error}, and it is always present whenever you work with non-integer values.

This is not a disaster --- the errors are typically tiny, on the order of $10^{-16}$ for the standard 64-bit format you will meet in the next section.
But they accumulate.
Add up a million slightly-wrong numbers and the total error can become large enough to matter.
In statistics, where you routinely sum thousands of data values or iterate an optimisation algorithm hundreds of times, this is not a theoretical concern.
It is a practical one.

\begin{Exercise}[title={Spot the accumulation}, label={ex:fp-accumulation}]
  A student adds $0.1$ to a running total $10{,}000$ times and expects the final result to be exactly $1000.0$.
  \begin{enumerate}
    \item Why might the result not be exactly $1000.0$?
    \item Would you expect the error to be larger or smaller than adding $0.1$ just once?
      Explain your reasoning in one or two sentences.
    \item In a statistical context, give one example where repeated addition of this kind might occur.
  \end{enumerate}
\end{Exercise}
\begin{Answer}[ref={ex:fp-accumulation}]
  \begin{enumerate}
    \item Because $0.1$ cannot be represented exactly in binary, each addition uses a slightly wrong value.
      The errors from each step add up rather than cancelling out perfectly.
    \item Larger, because each of the $10{,}000$ additions contributes a small error in the same direction.
      Errors tend to accumulate rather than cancel.
    \item Computing the sum $\sum x_i$ for a large dataset (e.g., summing $10{,}000$ exam scores to find the mean).
  \end{enumerate}
\end{Answer}

\subsection{Why This Belongs in a Statistics Course}

You might be wondering whether this is really a statistics problem.
It is.

Every time your software computes a mean, fits a regression line, or reports a $p$-value, it is doing arithmetic on floating-point numbers.
The results you read off the screen are already the product of thousands of these tiny approximations.
Most of the time the answers are reliable --- but knowing when to trust them, and when to be suspicious, requires understanding where the errors come from.

This chapter builds that understanding from the ground up.
We start with how numbers are stored (the IEEE 754 standard), then look at how errors arise and compound, and finish with practical strategies for writing statistical code that does not silently mislead you.

\begin{Exercise}[title={In your own words}, label={ex:fp-own-words}]
  Without looking back at the section, answer the following in two or three sentences:
  why can a computer not store the number $0.1$ exactly, and what does it do instead?
\end{Exercise}
\begin{Answer}[ref={ex:fp-own-words}]
  The number $0.1$ requires an infinitely repeating binary fraction, just as $\tfrac{1}{3}$ requires an infinitely repeating decimal.
  Because a computer has a finite number of bits, it truncates the binary expansion and stores the closest representable value.
  The tiny difference between $0.1$ and its stored approximation is the source of floating-point error.
\end{Answer}

\subsection*{Summary}
\begin{itemize}
  \item Computers store numbers in base 2 (binary), not base 10.
  \item Only fractions whose denominator is a power of 2 can be represented exactly in binary.
  \item Common decimal fractions like $0.1$ expand into infinite binary fractions and must be approximated.
  \item The approximation error is called floating-point error; it is tiny but accumulates over many operations.
  \item This matters in statistics because most computations involve many arithmetic steps on real-valued data.
\end{itemize}

% Generated without source material
% ============================================================
%  Numerical Methods in AP Statistics
%  Chapter: Floating Point Arithmetic & IEEE Standards
%  Section 2: Binary Representation of Numbers
% ============================================================
% Generated without source material

\section{Binary Representation of Numbers}

You already know that computers count in base~2.
Now let us make that concrete.
Before we can talk about how a computer stores a number like $-6.75$, we need
to understand how it stores the pieces: the sign, the whole part, and the
fractional part.
By the end of this section you will be able to convert any small decimal number
into its binary form by hand --- and more importantly, you will see exactly
where the limits of that representation come from.

%------------------------------------------------------------
\subsection{Whole Numbers in Binary}

Take the number $13$.
To convert it to binary, divide repeatedly by $2$ and record the remainders,
then read them from bottom to top.

\begin{center}
\begin{tabular}{rrl}
  \hline
  Division       & Quotient & Remainder \\
  \hline
  $13 \div 2$    & $6$      & $1$ \\
  $6  \div 2$    & $3$      & $0$ \\
  $3  \div 2$    & $1$      & $1$ \\
  $1  \div 2$    & $0$      & $1$ \quad $\leftarrow$ most significant bit \\
  \hline
\end{tabular}
\end{center}

Reading remainders from bottom to top: $13 = 1101_2$.
Check:
\[
  1 \times 2^3 + 1 \times 2^2 + 0 \times 2^1 + 1 \times 2^0
  \;=\; 8 + 4 + 0 + 1 \;=\; 13 \quad\checkmark
\]

The general rule: divide by $2$, record the remainder ($0$ or $1$), repeat
on the quotient until the quotient reaches $0$, then read remainders
upward.

\begin{Exercise}[title={Convert whole numbers to binary}, label={ex:bin-whole}]
  Convert each decimal integer to binary using the repeated division method.
  Show your working in the space provided, then write the binary result.

  \begin{center}
  \begin{tabular}{lll}
    \hline
    Decimal & Division working & Binary result \\
    \hline
    $7$     & & \\[24pt]
    $10$    & & \\[24pt]
    $25$    & & \\[24pt]
    $100$   & & \\[24pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:bin-whole}]
  \begin{itemize}
    \item $7 = 111_2$ \quad ($4 + 2 + 1$)
    \item $10 = 1010_2$ \quad ($8 + 2$)
    \item $25 = 11001_2$ \quad ($16 + 8 + 1$)
    \item $100 = 1100100_2$ \quad ($64 + 32 + 4$)
  \end{itemize}
\end{Answer}

You can also go in reverse --- convert a binary number back to decimal by
summing the powers of $2$ wherever a $1$ appears.

\begin{Exercise}[title={Binary back to decimal}, label={ex:bin-back}]
  Convert each binary number to decimal.
  Write out the power-of-2 sum to show your reasoning.

  \begin{center}
  \begin{tabular}{lll}
    \hline
    Binary       & Power-of-2 sum & Decimal \\
    \hline
    $1010_2$     & & \\[8pt]
    $11011_2$    & & \\[8pt]
    $101010_2$   & & \\[8pt]
    $11111111_2$ & & \\[8pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:bin-back}]
  \begin{itemize}
    \item $1010_2 = 8 + 2 = 10$
    \item $11011_2 = 16 + 8 + 2 + 1 = 27$
    \item $101010_2 = 32 + 8 + 2 = 42$
    \item $11111111_2 = 128 + 64 + 32 + 16 + 8 + 4 + 2 + 1 = 255 = 2^8 - 1$
  \end{itemize}
\end{Answer}

%------------------------------------------------------------
\subsection{Fractions in Binary}

Converting a fractional part works the other way: multiply by $2$ and record
the integer parts, reading them from top to bottom.

Take $0.625$:

\begin{center}
\begin{tabular}{rrl}
  \hline
  Multiplication    & Result & Integer part (bit) \\
  \hline
  $0.625 \times 2$  & $1.25$ & $1$ \\
  $0.25  \times 2$  & $0.5$  & $0$ \\
  $0.5   \times 2$  & $1.0$  & $1$ \quad $\leftarrow$ terminates \\
  \hline
\end{tabular}
\end{center}

Reading top to bottom: $0.625 = 0.101_2$.
Check:
\[
  1 \times 2^{-1} + 0 \times 2^{-2} + 1 \times 2^{-3}
  \;=\; 0.5 + 0 + 0.125 \;=\; 0.625 \quad\checkmark
\]

Now try $0.1$:

\begin{center}
\begin{tabular}{rrl}
  \hline
  Multiplication  & Result & Integer part (bit) \\
  \hline
  $0.1 \times 2$  & $0.2$  & $0$ \\
  $0.2 \times 2$  & $0.4$  & $0$ \\
  $0.4 \times 2$  & $0.8$  & $0$ \\
  $0.8 \times 2$  & $1.6$  & $1$ \\
  $0.6 \times 2$  & $1.2$  & $1$ \\
  $0.2 \times 2$  & $0.4$  & $0$ \quad $\leftarrow$ same as row 2; repeats forever \\
  \hline
\end{tabular}
\end{center}

The fractional part cycles back to $0.2$ and the pattern $0011$ repeats
without end:
\[
  0.1_{10} \;=\; 0.0\overline{0011}_2
\]
There is no finite binary string for $0.1$.
This is the root cause of the \texttt{0.1 + 0.2} result you saw in Section~1.

The rule that tells you in advance whether a fraction will terminate is
simple: a decimal fraction terminates in binary if and only if its denominator
(in lowest terms) is a power of $2$.

\begin{center}
\begin{tabular}{llll}
  \hline
  Fraction        & Denominator & Power of 2? & Binary \\
  \hline
  $\frac{1}{2}$   & $2 = 2^1$   & Yes & $0.1_2$ (terminates) \\
  $\frac{3}{8}$   & $8 = 2^3$   & Yes & $0.011_2$ (terminates) \\
  $\frac{1}{5}$   & $5$         & No  & $0.\overline{0011}_2$ (repeats) \\
  $\frac{1}{10}$  & $10 = 2 \times 5$ & No & $0.0\overline{0011}_2$ (repeats) \\
  \hline
\end{tabular}
\end{center}

\begin{Exercise}[title={Terminate or repeat?}, label={ex:bin-frac}]
  Apply the repeated multiplication method to each fraction.
  Stop after 8 steps or when the process terminates, whichever comes first.
  State whether the binary expansion is exact or repeating.

  \begin{center}
  \begin{tabular}{llll}
    \hline
    Fraction & Denominator a power of 2? & Binary expansion & Exact or repeating? \\
    \hline
    $0.5$    & & & \\[8pt]
    $0.75$   & & & \\[8pt]
    $0.2$    & & & \\[8pt]
    $0.125$  & & & \\[8pt]
    $0.3$    & & & \\[8pt]
    $0.4375$ & & & \\[8pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:bin-frac}]
  \begin{itemize}
    \item $0.5 = \frac{1}{2}$: denominator $2^1$ $\Rightarrow$ exact.
          $0.5 = 0.1_2$.
    \item $0.75 = \frac{3}{4}$: denominator $2^2$ $\Rightarrow$ exact.
          $0.75 = 0.11_2$.
    \item $0.2 = \frac{1}{5}$: denominator $5$ $\Rightarrow$ repeating.
          $0.2 = 0.0\overline{0011}_2$.
    \item $0.125 = \frac{1}{8}$: denominator $2^3$ $\Rightarrow$ exact.
          $0.125 = 0.001_2$.
    \item $0.3 = \frac{3}{10}$: denominator $10 = 2 \times 5$
          $\Rightarrow$ repeating.
          $0.3 = 0.0\overline{1001}_2$.
    \item $0.4375 = \frac{7}{16}$: denominator $2^4$ $\Rightarrow$ exact.
          $0.4375 = 0.0111_2$.
  \end{itemize}
\end{Answer}

%------------------------------------------------------------
\subsection{Combining Whole and Fractional Parts}

To represent a number like $6.75$ in binary, convert the integer part and
the fractional part separately, then join them at the binary point.

\begin{center}
\begin{tabular}{ll}
  \hline
  Part             & Conversion \\
  \hline
  Integer: $6$     & $6 \div 2 = 3\,R\,0$;\; $3 \div 2 = 1\,R\,1$;\;
                     $1 \div 2 = 0\,R\,1$ $\;\Rightarrow\; 110_2$ \\
  Fractional: $0.75$ & $0.75 \times 2 = 1.5$ (bit $1$);\;
                       $0.5 \times 2 = 1.0$ (bit $1$) $\;\Rightarrow\; 0.11_2$ \\
  Combined         & $6.75 = 110.11_2$ \\
  \hline
\end{tabular}
\end{center}

Check: $4 + 2 + 0.5 + 0.25 = 6.75$. $\checkmark$

\begin{Exercise}[title={Convert mixed numbers to binary}, label={ex:bin-mixed}]
  Convert each number to binary, working the integer and fractional parts
  separately.

  \begin{center}
  \begin{tabular}{llll}
    \hline
    Number   & Integer part (binary) & Fractional part (binary) & Full binary \\
    \hline
    $5.5$    & & & \\[8pt]
    $3.25$   & & & \\[8pt]
    $12.125$ & & & \\[8pt]
    $0.375$  & & & \\[8pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:bin-mixed}]
  \begin{itemize}
    \item $5.5 = 101.1_2$ \quad ($4 + 1 + 0.5$)
    \item $3.25 = 11.01_2$ \quad ($2 + 1 + 0.25$)
    \item $12.125 = 1100.001_2$ \quad ($8 + 4 + 0.125$)
    \item $0.375 = 0.011_2$ \quad ($0.25 + 0.125$)
  \end{itemize}
\end{Answer}

%------------------------------------------------------------
\subsection{Negative Numbers and the Sign Bit}

So far all our numbers have been positive.
Computers reserve one bit --- always the leftmost --- to encode the sign.
When the \newterm{sign bit}\index{sign bit} is $0$ the number is positive;
when it is $1$ the number is negative.

A complete binary number therefore has three parts:

\begin{itemize}
  \item \newterm{Sign bit}\index{sign bit}: $0$ for positive, $1$ for negative.
  \item Integer part: the whole-number digits in base~$2$.
  \item Fractional part: the digits after the binary point.
\end{itemize}

So $-6.75$ in binary is written $-110.11_2$ and stored with a sign bit of $1$.

\subsubsection{Why Not Just Flip the Sign Bit?}

You might think storing $-13$ is as simple as taking $1101_2$ and changing the
sign bit to $1$.
This scheme is called \newterm{sign-magnitude}\index{sign-magnitude} and it
does work --- but it has a quirk: it produces two representations of zero
($+0$ and $-0$), which complicates arithmetic circuits.

Modern hardware uses a cleverer scheme called
\newterm{two's complement}\index{two's complement} for integers.
In two's complement, negating a number means flipping every bit and adding
$1$.

\begin{center}
\begin{tabular}{lll}
  \hline
  Step & Example: find $-13$ in 8-bit two's complement & Bit pattern \\
  \hline
  Start with $+13$        & $13 = 1101_2$                    & \texttt{00001101} \\
  Flip every bit           & $0$ becomes $1$, $1$ becomes $0$ & \texttt{11110010} \\
  Add $1$                  & $\texttt{11110010} + 1$          & \texttt{11110011} \\
  \hline
  Result: $-13$            &                                   & \texttt{11110011} \\
  \hline
\end{tabular}
\end{center}

Check: $\texttt{00001101} + \texttt{11110011} = \texttt{100000000}$.
The leading $1$ overflows an 8-bit register and is discarded, leaving
$\texttt{00000000} = 0$. \checkmark

Two's complement is used for integer storage in virtually every modern
processor.
Floating-point numbers use a different (but related) trick for the exponent ---
biased encoding --- which you will see in Section~3.

\begin{Exercise}[title={Two's complement}, label={ex:bin-twos}]
  \begin{enumerate}
    \item Find the 8-bit two's complement representation of $-7$.
          Show the flip-and-add-1 steps.
    \item Find the 8-bit two's complement representation of $-1$.
          What pattern do you notice?
    \item What is the largest positive integer representable in 8-bit
          two's complement?
          What is the most negative?
    \item An 8-bit two's complement number reads \texttt{10000000}.
          What decimal value does it represent?
          (Hint: apply flip-and-add-1 to find the magnitude, then check the
          sign bit.)
  \end{enumerate}
\end{Exercise}
\begin{Answer}[ref={ex:bin-twos}]
  \begin{enumerate}
    \item $+7 = \texttt{00000111}$;\; flip: \texttt{11111000};\;
          add 1: \texttt{11111001}.\; So $-7 = \texttt{11111001}$.
    \item $+1 = \texttt{00000001}$;\; flip: \texttt{11111110};\;
          add 1: \texttt{11111111}.\; So $-1 = \texttt{11111111}$ --- all ones.
    \item Largest positive: \texttt{01111111} $= 127$.
          Most negative: \texttt{10000000} $= -128$.
          (Note the asymmetry: there is no $+128$ in 8-bit two's complement.)
    \item \texttt{10000000}: sign bit is $1$ (negative).
          Flip: \texttt{01111111};\; add 1: \texttt{10000000}.
          The magnitude wraps --- this is the special case $-128$.
  \end{enumerate}
\end{Answer}

%------------------------------------------------------------
\subsection{Putting It All Together}

A full binary number --- sign, integer part, fractional part --- is now within
reach.

\begin{Exercise}[title={Full binary representation}, label={ex:bin-full}]
  Convert each number to binary, filling in the sign bit, integer part, and
  fractional part separately.

  \begin{center}
  \begin{tabular}{lllll}
    \hline
    Number    & Sign bit & Integer part (binary) & Fractional part (binary)
              & Full binary \\
    \hline
    $+5.5$    & & & & \\[8pt]
    $-3.25$   & & & & \\[8pt]
    $+0.375$  & & & & \\[8pt]
    $-12.125$ & & & & \\[8pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:bin-full}]
  \begin{itemize}
    \item $+5.5$: sign $0$,\; integer $101_2$,\; fraction $.1_2$
          $\;\Rightarrow\; 0 \mid 101.1_2$
    \item $-3.25$: sign $1$,\; integer $11_2$,\; fraction $.01_2$
          $\;\Rightarrow\; 1 \mid 11.01_2$
    \item $+0.375$: sign $0$,\; integer $0$,\; fraction $.011_2$
          $\;\Rightarrow\; 0 \mid 0.011_2$
    \item $-12.125$: sign $1$,\; integer $1100_2$,\; fraction $.001_2$
          $\;\Rightarrow\; 1 \mid 1100.001_2$
  \end{itemize}
\end{Answer}

%------------------------------------------------------------
\subsection*{Summary}
\begin{itemize}
  \item To convert a whole number to binary: divide repeatedly by $2$
        and read remainders upward.
  \item To convert a fraction to binary: multiply repeatedly by $2$
        and read integer parts downward.
  \item A fraction terminates in binary if and only if its denominator
        (in lowest terms) is a power of $2$; otherwise it repeats indefinitely.
  \item Whole and fractional binary parts are joined at the binary point:
        $6.75 = 110.11_2$.
  \item The sign bit ($0$ = positive, $1$ = negative) is stored separately
        as the leftmost bit.
  \item Modern processors use two's complement for integer negation;
        floating-point formats use biased encoding for the exponent instead
        (covered in Section~3).
\end{itemize}

% Generated without source material
\section{The IEEE 754 Standard}


Every time your calculator, phone, or laptop does arithmetic with a decimal
number, it is following a rulebook.
That rulebook is called \newterm{IEEE 754}\index{IEEE 754}, published in 1985 by
the Institute of Electrical and Electronics Engineers.
Before it existed, different computers stored numbers in incompatible ways --- a
program that ran correctly on one machine might give wrong answers on another.
IEEE 754 fixed that by giving every manufacturer the same blueprint.

In this section we open that blueprint and read it carefully.
By the end you will know which bits store which part of a number, why the format
is cleverer than it first looks, what its precision limits are, and how to
decode a raw 64-bit string by hand.

%------------------------------------------------------------
\subsection{The Core Idea: Scientific Notation in Binary}

You have used scientific notation before.
The number $-0.00675$ in scientific notation is $-6.75 \times 10^{-3}$.
Pull out the significant digits ($6.75$), record the scale ($10^{-3}$), note
the sign ($-$).
Three pieces, no wasted zeros.

IEEE 754 does exactly the same thing --- in binary.
The number $-6.75$ in binary is $-110.11_2$.
Shifting the binary point left until exactly one non-zero digit sits to its
left gives:
\[
  -6.75 \;=\; -1.1011_2 \;\times\; 2^{2}
\]

That leading digit is always $1$ in binary (the only non-zero binary digit
\emph{is} $1$), so the computer never needs to store it.
This is called the \newterm{implicit leading one}\index{implicit leading one},
and it gives one free bit of precision at no cost.

The three fields IEEE 754 stores are:

\begin{itemize}
  \item \newterm{Sign}\index{sign bit} --- one bit: $0$ for positive,
        $1$ for negative.
  \item \newterm{Exponent}\index{exponent} --- the power of $2$ that scales
        the number.
  \item \newterm{Mantissa}\index{mantissa} (also called the
        \newterm{significand}\index{significand}) --- the binary digits after
        the implicit leading $1$.
\end{itemize}

\begin{Exercise}[title={Normalise to binary scientific notation},
                 label={ex:ieee-normalise}]
  Write each number in the form $\pm 1.\mathrm{mantissa}_2 \times 2^{e}$.
  State the sign bit, exponent $e$, and the mantissa bits.

  \begin{center}
  \begin{tabular}{lllll}
    \hline
    Number   & Binary        & Normalised form & Exponent $e$ & Mantissa bits \\
    \hline
    $+13.0$  & $1101.0_2$    & & & \\[6pt]
    $+0.625$ & $0.101_2$     & & & \\[6pt]
    $-3.75$  & $-11.11_2$    & & & \\[6pt]
    $+1.0$   & $1.0_2$       & & & \\[6pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:ieee-normalise}]
  $+13.0$: $1.1010_2 \times 2^3$;\; sign $0$;\; $e=3$;\; mantissa $=1010$.\\
  $+0.625$: $1.01_2 \times 2^{-1}$;\; sign $0$;\; $e=-1$;\; mantissa $=01$.\\
  $-3.75$: $-1.111_2 \times 2^{1}$;\; sign $1$;\; $e=1$;\; mantissa $=111$.\\
  $+1.0$: $1.0_2 \times 2^{0}$;\; sign $0$;\; $e=0$;\; mantissa $=0$
  (nothing after the binary point).
\end{Answer}

%------------------------------------------------------------
\subsection{Single Precision vs.\ Double Precision}

IEEE 754 defines two main formats.

\begin{center}
\begin{tabular}{lrrrr}
  \hline
  Format                    & Total bits & Sign & Exponent bits & Mantissa bits \\
  \hline
  Single precision (32-bit) & 32         & 1    & 8             & 23            \\
  Double precision (64-bit) & 64         & 1    & 11            & 52            \\
  \hline
\end{tabular}
\end{center}

Here is how the 64 bits are laid out in a double-precision number:

\begin{center}
\begin{tabular}{|c|c|c|}
  \hline
  \texttt{S} \;(1 bit)
  & \texttt{EEEEEEEEEEE} \;(11 bits)
  & \texttt{MMMM\ldots M} \;(52 bits) \\
  sign & exponent & mantissa \\
  \hline
\end{tabular}
\end{center}

% [figure placeholder: colour-coded 64-bit memory diagram, sign in red,
%  exponent in blue, mantissa in green, bit positions labelled 63 down to 0]

Python, R, and virtually all statistical software use double precision by
default.
When your code writes \texttt{3.14}, the computer allocates all 64 bits and
fills them with the closest representable approximation to $\pi$.

\begin{Exercise}[title={How many numbers can we represent?},
                 label={ex:ieee-count}]
  \begin{enumerate}
    \item A single-precision number uses 32 bits, each either $0$ or $1$.
          How many distinct 32-bit patterns exist?
    \item A double-precision number uses 64 bits.
          How many distinct patterns does that allow?
          Give your answer as a power of $2$ and as an approximate decimal.
    \item What does the computer do when a real number falls exactly halfway
          between two representable double-precision values?
    \item Would you expect to run out of representable values in a typical
          statistics calculation? Justify your answer briefly.
  \end{enumerate}
\end{Exercise}
\begin{Answer}[ref={ex:ieee-count}]
  \begin{enumerate}
    \item $2^{32} = 4{,}294{,}967{,}296$ --- about 4.3 billion patterns.
    \item $2^{64} \approx 1.8 \times 10^{19}$ --- about 18 quintillion
          patterns.
    \item The computer rounds to the nearest representable value.
          For ties, it applies \emph{round half to even} (banker's rounding),
          choosing the value whose last mantissa bit is $0$.
    \item In practice, no.
          The largest finite double-precision number is about $1.8 \times
          10^{308}$; statistical values rarely approach this.
          Loss of \emph{precision}, not overflow, is the real concern.
  \end{enumerate}
\end{Answer}

%------------------------------------------------------------
\subsection{Biased Encoding of the Exponent}

The exponent must represent both positive and negative powers of $2$, yet no
sign bit has been reserved for it.
How does the computer distinguish $2^{-10}$ from $2^{10}$?

The answer is \newterm{biased encoding}\index{biased encoding}, also called
\emph{excess-$N$} or \emph{offset binary}.
Instead of storing the actual exponent $e$, the computer stores $e +
\text{bias}$, which is always a non-negative integer.

\[
  \text{stored value} = e + \text{bias}
  \qquad\Longleftrightarrow\qquad
  e = \text{stored value} - \text{bias}
\]

The bias is $\mathbf{1023}$ for double precision and $\mathbf{127}$ for single
precision.

\begin{center}
\begin{tabular}{rrl}
  \hline
  Actual exponent $e$ & Stored value ($e + 1023$) & 11-bit binary \\
  \hline
  $-1023$ & $0$    & \texttt{00000000000}\quad (reserved: zero and subnormals)\\
  $-10$   & $1013$ & \texttt{01111110101} \\
  $-1$    & $1022$ & \texttt{01111111110} \\
  $0$     & $1023$ & \texttt{01111111111} \\
  $1$     & $1024$ & \texttt{10000000000} \\
  $10$    & $1033$ & \texttt{10000001001} \\
  $1023$  & $2046$ & \texttt{11111111110} \\
  $1024$  & $2047$ & \texttt{11111111111}\quad (reserved: $\pm\infty$ and NaN)\\
  \hline
\end{tabular}
\end{center}

Stored values $0$ and $2047$ are reserved for special numbers --- zero,
subnormals, $\pm\infty$, and NaN --- which you will meet in Section~5.
Every stored value from $1$ to $2046$ represents a normal, valid exponent.

Why use this trick?
With biased encoding the exponent field can be compared directly as an ordinary
unsigned integer: a larger stored value always means a larger magnitude.
No special logic for negative exponents is needed in hardware, making
floating-point comparison very fast.

\begin{Exercise}[title={Decode and encode the exponent}, label={ex:ieee-exp}]

  \textbf{Part A --- Decoding (single precision, bias $= 127$).}
  Find the actual exponent $e$ and the power of $2$ it represents.

  \begin{center}
  \begin{tabular}{lll}
    \hline
    Stored value & Actual exponent $e$ & Power of 2 \\
    \hline
    $127$ & & \\[6pt]
    $128$ & & \\[6pt]
    $126$ & & \\[6pt]
    $1$   & & \\[6pt]
    $254$ & & \\[6pt]
    \hline
  \end{tabular}
  \end{center}

  \textbf{Part B --- Encoding (double precision, bias $= 1023$).}
  Find the stored value and write it in 11-bit binary.

  \begin{center}
  \begin{tabular}{lll}
    \hline
    Actual exponent $e$ & Stored value & 11-bit binary \\
    \hline
    $0$   & & \\[6pt]
    $3$   & & \\[6pt]
    $-1$  & & \\[6pt]
    $-52$ & & \\[6pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:ieee-exp}]
  \textbf{Part A:}
  \begin{itemize}
    \item $127 - 127 = 0 \;\Rightarrow\; 2^0 = 1$
    \item $128 - 127 = 1 \;\Rightarrow\; 2^1 = 2$
    \item $126 - 127 = -1 \;\Rightarrow\; 2^{-1} = 0.5$
    \item $1 - 127 = -126 \;\Rightarrow\; 2^{-126} \approx 1.18 \times
          10^{-38}$ (near underflow for single precision)
    \item $254 - 127 = 127 \;\Rightarrow\; 2^{127} \approx 1.70 \times
          10^{38}$ (near overflow for single precision)
  \end{itemize}

  \textbf{Part B:}
  \begin{itemize}
    \item $e = 0$: \; stored $= 1023 = \texttt{01111111111}_2$
    \item $e = 3$: \; stored $= 1026 = \texttt{10000000010}_2$
    \item $e = -1$: \; stored $= 1022 = \texttt{01111111110}_2$
    \item $e = -52$: \; stored $= 971 = \texttt{01111001011}_2$
  \end{itemize}
\end{Answer}

%------------------------------------------------------------
\subsection{The Mantissa in Detail}

The mantissa stores the binary digits after the implicit leading $1$.
For double precision there are 52 of them; together with the implicit $1$ you
have 53 significant binary digits.

Let us encode $+13.0$ step by step.

\medskip
\textbf{Step 1 --- Convert to binary.}
\[
  13 = 8 + 4 + 1 = 2^3 + 2^2 + 2^0 = 1101_2
  \qquad\Longrightarrow\qquad
  +13.0 = 1101.0_2
\]

\textbf{Step 2 --- Normalise.}
Shift the binary point left by $3$ places so one digit sits to its left:
\[
  1101.0_2 \;=\; 1.1010_2 \times 2^3
\]
Exponent $e = 3$. Mantissa bits (after the leading $1$): \texttt{1010}.

\textbf{Step 3 --- Sign bit.}
Positive $\Rightarrow$ sign bit $= 0$.

\textbf{Step 4 --- Stored exponent.}
\[
  3 + 1023 = 1026 \;=\; 2^{10} + 2^1 \;=\; \texttt{10000000010}_2
\]

\textbf{Step 5 --- Mantissa field.}
The significand after the implicit $1$ is \texttt{1010}.
Pad to 52 bits with trailing zeros.

\medskip
\textbf{Full 64-bit double-precision string for $+13.0$:}
\[
\underbrace{0}_{S}
\quad
\underbrace{\texttt{10000000010}}_{E\;(11\;\mathrm{bits})}
\quad
\underbrace{\texttt{1010}\overbrace{\texttt{00\!\cdots\!0}}^{48\;\mathrm{zeros}}}_{M\;(52\;\mathrm{bits})}
\]

Because $13.0$ is a whole number whose binary expansion terminates, it is
stored exactly.
Try the same process with $0.1$ and the mantissa bits never settle ---
the 52-bit string is an approximation.

\begin{Exercise}[title={Encode a number by hand}, label={ex:ieee-encode}]
  Encode $+6.5$ in double-precision IEEE 754. Show every step.

  \begin{center}
  \begin{tabular}{ll}
    \hline
    Step & Your answer \\
    \hline
    1.\; Binary form of $6.5$ & \\[8pt]
    2.\; Normalised form
         $\;1.\underline{\hspace{2.2cm}}_2 \times 2^{\,\square}$ & \\[8pt]
    3.\; Sign bit & \\[8pt]
    4.\; Stored exponent $= e + 1023$ & \\[8pt]
    5.\; Stored exponent in 11-bit binary & \\[8pt]
    6.\; Mantissa bits (first 8, then describe padding) & \\[8pt]
    7.\; Full 64-bit string
         \;$S \mid E_{11} \mid M_{52}$ & \\[8pt]
    \hline
  \end{tabular}
  \end{center}
\end{Exercise}
\begin{Answer}[ref={ex:ieee-encode}]
  \begin{enumerate}
    \item $6.5 = 110.1_2$
    \item $1.101_2 \times 2^2$;\; $e = 2$
    \item Sign bit $= 0$
    \item Stored exponent $= 2 + 1023 = 1025$
    \item $1025 = 2^{10} + 2^0 = \texttt{10000000001}_2$
    \item Mantissa: \texttt{1010} then 48 zeros (first 8 bits: \texttt{10100000})
    \item \texttt{0 | 10000000001 | 1010000000000000000000000000000000000000000000000000}
  \end{enumerate}
\end{Answer}

\begin{Exercise}[title={Decode a 64-bit string}, label={ex:ieee-decode}]
  The following 64-bit double-precision string is given:

  \medskip
  \begin{center}
  \texttt{0 \; 10000000000 \;
          1000000000000000000000000000000000000000000000000000}
  \end{center}
  \medskip

  \begin{enumerate}
    \item What is the sign?
    \item Convert the 11-bit exponent field to a decimal stored value.
    \item What is the actual exponent $e$?
    \item Write the full significand, including the implicit leading $1$.
    \item What decimal number does this bit string represent?
  \end{enumerate}
\end{Exercise}
\begin{Answer}[ref={ex:ieee-decode}]
  \begin{enumerate}
    \item Sign bit $= 0$ --- positive.
    \item $\texttt{10000000000}_2 = 2^{10} = 1024$.
    \item $e = 1024 - 1023 = 1$.
    \item Significand $= 1.1_2$ (implicit $1$, then mantissa bit $1$, rest
          zeros).
    \item $1.1_2 \times 2^1 = 11.0_2 = 3.0$.\;
          The string represents $+3.0$.
  \end{enumerate}
\end{Answer}

%------------------------------------------------------------
\subsection{Precision and Range}

\subsubsection{Range}

The largest finite double-precision number has stored exponent $2046$
(actual $e = 1023$) and all mantissa bits set to $1$:
\[
  (2 - 2^{-52}) \times 2^{1023} \;\approx\; 1.8 \times 10^{308}
\]
The estimated number of atoms in the observable universe is about $10^{80}$,
well inside this range.
The smallest positive \emph{normal} double-precision number is about
$2.2 \times 10^{-308}$.
In statistics, overflow is almost never the problem; underflow is somewhat more
likely but still rare.

\subsubsection{Precision and Machine Epsilon}

With 53 effective significand bits, double precision provides:
\[
  53 \times \log_{10}(2) \;\approx\; 53 \times 0.301 \;\approx\; 15.95
  \quad\text{significant decimal digits.}
\]

The fundamental precision limit is called \newterm{machine epsilon}\index{machine epsilon},
written $\varepsilon_{\mathrm{mach}}$.
It is defined as the smallest positive number such that
\[
  \mathrm{fl}(1.0 + \varepsilon_{\mathrm{mach}}) \neq 1.0
\]
where $\mathrm{fl}(\cdot)$ denotes the floating-point result of an operation.
For double precision:
\[
  \varepsilon_{\mathrm{mach}} = 2^{-52} \;\approx\; 2.22 \times 10^{-16}
\]

Two important clarifications:
\begin{itemize}
  \item $\varepsilon_{\mathrm{mach}}$ is \emph{not} the smallest representable
        positive number.
        Numbers far smaller than $\varepsilon_{\mathrm{mach}}$ exist
        (subnormals go down to about $5 \times 10^{-324}$).
  \item $\varepsilon_{\mathrm{mach}}$ is the worst-case \emph{relative} gap
        near $1.0$.
        Every correctly rounded floating-point operation introduces a relative
        error of at most $\tfrac{1}{2}\varepsilon_{\mathrm{mach}}$.
\end{itemize}

You will compute $\varepsilon_{\mathrm{mach}}$ experimentally in Section~4.

\begin{Exercise}[title={Precision in practice}, label={ex:ieee-precision}]
  \begin{enumerate}
    \item A student stores $\pi \approx 3.14159265358979323846\ldots$ as a
          double-precision float.
          How many of those decimal digits are reliable?

    \item She then computes $\pi - 3.14159265358979$.
          The mathematically correct result is approximately
          $3.23 \times 10^{-15}$.
          Should she trust all digits of the floating-point result?
          Why or why not?

    \item Single precision has 23 mantissa bits.
          Use the approximation $2^{10} \approx 10^3$ to estimate how many
          significant decimal digits it provides.
          Why might this matter when choosing a numeric data type?

    \item The true value of $0.1$ is
          $0.1000000000000000055511151231\ldots$ when stored in double
          precision.
          Compute the absolute error and express it as a multiple of
          $\varepsilon_{\mathrm{mach}} \approx 2.22 \times 10^{-16}$.
  \end{enumerate}
\end{Exercise}
\begin{Answer}[ref={ex:ieee-precision}]
  \begin{enumerate}
    \item About 15--16 digits.
          The stored double-precision value of $\pi$ agrees with the true
          value through the 15th decimal digit; the 16th is uncertain.

    \item No.
          Subtracting two nearly equal numbers causes
          \newterm{catastrophic cancellation}\index{catastrophic cancellation}:
          the significant digits cancel and the result is dominated by
          rounding error.
          Only 1--2 digits of the answer are meaningful.
          Section~6 treats this in full.

    \item $23 \times \frac{3}{10} \approx 7$ significant decimal digits.
          Single precision is common in graphics and machine learning where
          memory and speed matter more than numerical accuracy, but 7 digits
          is too coarse for most statistical computation.

    \item Absolute error $\approx 5.55 \times 10^{-18}$.
          Dividing by $\varepsilon_{\mathrm{mach}} \approx 2.22 \times
          10^{-16}$ gives $\approx 0.025 \,\varepsilon_{\mathrm{mach}}$ ---
          well within the guaranteed half-epsilon bound.
  \end{enumerate}
\end{Answer}

%------------------------------------------------------------
\subsection*{Summary}
\begin{itemize}
  \item IEEE 754 (1985) is the universal standard for storing floating-point
        numbers, adopted by every modern processor.
  \item Every number is stored as
        $\pm\, 1.\mathrm{mantissa}_2 \times 2^e$.
        The leading $1$ is implicit and never stored, giving one free bit.
  \item Double precision: 1 sign $+$ 11 exponent $+$ 52 mantissa $= 64$
        bits total.
  \item The exponent is stored with bias $1023$; values $0$ and $2047$ are
        reserved for special cases.
  \item 53 effective significand bits $\Rightarrow$ roughly 15--16 significant
        decimal digits.
  \item Machine epsilon $\varepsilon_{\mathrm{mach}} = 2^{-52}
        \approx 2.22 \times 10^{-16}$ is the relative precision limit.
  \item In statistical computing, precision is almost always the concern,
        not range.
\end{itemize}

% Generated without source material